$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-05-07 qdpm_paper.tex
updated and added some figures
Worked on text
symbols in table

@@

@@

Over the last several decades, experimental and theoretical work has demonstrated that strongly interacting quantum many-body systems support exotic quantum phases phases of matter not described by the Landau symmetry breaking paradigm. One class of such quantum phases of matter are topologically ordered quantum liquids~\cite{Wen1990}, first realized experimentally via the fractional quantum Hall effect~\cite{FQHE????}. Despite the absence of conventional symmetry breaking and a local order parameter, topological quantum liquids possess a non-trivial quantum order that distinguishes them from trivial liquid phases~\cite{Nayak2008}. Although not detectable by local operators, topological order arises due to a non-local entanglement structure that leaves a signatures in the bipartite entanglement entropy~\cite{Levin2006a,Kitaev2006b}. Such topological phases may provide the basis for fault-tolerant quantum information processing~\cite{Freedman2001,Kitaev2003}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \begin{tabular}{| l | c | c | c |}
  	\hline
      	type & $n_v$ & $Q_v^{\rm{QDM}}$ & $Q_v^{\rm{FPLM}}$  \\
	\hline
	monomer & $0$ & -1 & -2 \\
	\hline
	dimer & $1$ & 0 & -1 \\
	\hline
	trimer & $2$ & 1 & 0 \\
	\hline
	tetramer & $3$ & 2 & 1 \\
	\hline
	pentamer & $4$ & 3 & 2 \\
	\hline
  \end{tabular}
  \caption{Defects where $n_v$ dimers touch a vertex and their associated $U(1)$ charges on the A sub-lattice in both the QDM and QFPLM. Defects on the B sub-lattice have opposite charge.}
  \label{tab:defects}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We now consider relaxing the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) (and the closely related even and odd sector Toric Codes). In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{TC}}$ is expanded accordingly \textcolor{red}{Owen: Add symbols for 4-0 resonance, and 1-3 resonances and 2-2 resonance}:
\begin{align}
\mathcal{C}_{\rm{TC}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{ \left(\vrtplqt,\hrzplqt\right),\left(\vrtplqt,\hrzplqt\right),\left(\vrtplqt,\hrzplqt\right),\left(\vrtplqt,\hrzplqt\right) \bigr\}
\end{align}
where we implicitly include rotational symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

@@

We explicitly demonstrate how the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the $U(1)$ topological winding number $\Phi_X$ histogram. This is shown in \figref{fig:u1_wind_qdpm}, where the different colors represent calculations that were initialized to different winding sectors. We confirm that $\Phi_{X,Y}$ are conserved modulo $3$, leading to $9$-fold topological degeneracy. Notice that $\Phi_X = \pm L/2$ correspond to ordered configurations, and therefore the probability of $\Phi_{XY}$ reaching these extrema vanishes in the RK state.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=0.8\linewidth]{u1_wind_qdpm.pdf}
    \caption{The histogram of the $U(1)$ winding number $\Phi_X$ as measured in the QDPM on a $16\times16$ lattice. Different colors correspond to calculations initialized to different winding sectors. Note that $\Phi_X$ is conserved modulo $3$, resulting in a $9$-fold topological degeneracy.}
    \label{fig:u1_wind_qdpm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}
        \centering
        \includegraphics[width=0.75\columnwidth]{dimer_gry_scale_qdpm.pdf}
        \caption{Gray scale of the dimer correlation function between the dimer in the upper
        left corner (colored red) and all other dimers on a $16\times16$ lattice. The 
         red dimers indicate correlations that are off the scale shown at the
        bottom of the figure. \textcolor{red}{Owen: make the origin dimer a different color that the off-scale dimers}.}
        \label{fig:dimer_gry_scale_qdpm}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spacial_dmr_cor.pdf}
    \caption{Correlation function between parallel dimers one the same (blue) and different (red) 
    sub-lattice on an $L=32$ lattice. The lines represent least-squares fits to exponential decay. }
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
   \begin{figure}[]
        \centering
        \includegraphics[width=0.75\columnwidth]{qdpm_dmr_HM_16x16.pdf}
        \caption{QDPM dimer structure factor $\gamma=0.5$.}
        \label{fig:spacial_z3vis_cor}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=0.75\columnwidth]{qdpm_vis_HM_16x16.pdf}
        \caption{QDPM $Z_3$ vison structure factor $\gamma=0.5$.}
        \label{fig:spacial_z3vis_cor}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\newcommand*{\fplqt}{%
    ]{Symbols/full_plqt} }

@@

\newcommand*{\threedownplqt}{%
    ]{Symbols/threedownplqt.pdf} }
\newcommand*{\threeupplqt}{%
    ]{Symbols/threeupplqt.pdf} }
\newcommand*{\threerightplqt}{%
    ]{Symbols/threerightplqt.pdf} }
\newcommand*{\threeleftplqt}{%
    ]{Symbols/threeleftplqt.pdf} }
\newcommand*{\onedownplqt}{%
    ]{Symbols/onedownplqt.pdf} }
\newcommand*{\oneupplqt}{%
    ]{Symbols/oneupplqt.pdf} }
\newcommand*{\onerightplqt}{%
    ]{Symbols/onerightplqt.pdf} }
\newcommand*{\oneleftplqt}{%
    ]{Symbols/oneleftplqt.pdf} }
\newcommand*{\twoupplqt}{%
    ]{Symbols/twoupplqt.pdf} }
\newcommand*{\twodownplqt}{%
    ]{Symbols/twodownplqt.pdf} }
\newcommand*{\svrtmon}{%
      height=3.5ex,% adjust to suit
    ]{Symbols/sing_vrt_monomer.pdf} }
\newcommand*{\svrtdmr}{%
      height=3.5ex,% adjust to suit
    ]{Symbols/sing_vrt_dimer.pdf} }
\newcommand*{\svrttrmr}{%
      height=3.5ex,% adjust to suit
    ]{Symbols/sing_vrt_trimer.pdf} }
\newcommand*{\svrtquat}{%
      height=3.5ex,% adjust to suit
    ]{Symbols/sing_vrt_quatramer.pdf} }
\newcommand*{\svrtpent}{%
      height=3.5ex,% adjust to suit
    ]{Symbols/sing_vrt_pentamer.pdf} }

@@

Over the last several decades, experimental and theoretical work has demonstrated that strongly interacting quantum many-body systems support exotic quantum phases of matter not described by the Landau symmetry breaking paradigm. One class of such quantum phases of matter are topologically ordered quantum liquids~\cite{Wen1990}, first realized experimentally via the fractional quantum Hall effect~\cite{FQHE????}. Despite the absence of conventional symmetry breaking and a local order parameter, topological quantum liquids possess a non-trivial quantum order that distinguishes them from trivial liquid phases~\cite{Nayak2008}. Although not detectable by local operators, topological order arises due to a non-local entanglement structure that leaves a signatures in the bipartite entanglement entropy~\cite{Levin2006a,Kitaev2006b}. Such topological phases may provide the basis for fault-tolerant quantum information processing~\cite{Freedman2001,Kitaev2003}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \begin{tabular}{| l | c | c | c | c |}
  	\hline
    type & $n_v$ & $Q_v^{\rm{QDM}}$ & $Q_v^{\rm{FPLM}}$  &  \\
	\hline
    monomer & $0$ & -1 & -2 & $\svrtmon$ \\
	\hline
    dimer & $1$ & 0 & -1 &  $\svrtdmr$ \\
	\hline
    trimer & $2$ & 1 & 0 &  $\svrttrmr $ \\
	\hline
    tetramer & $3$ & 2 & 1 &  $\svrtquat$ \\
	\hline
    pentamer & $4$ & 3 & 2 &  $\svrtpent$ \\
	\hline
  \end{tabular}
  \caption{Defects where $n_v$ dimers touch a vertex and their associated $U(1)$ charges on the A sub-lattice in both the QDM and QFPLM. Defects on the B sub-lattice have opposite charge.}
  \label{tab:defects}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We now consider relaxing the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) (and the closely related even and odd sector Toric Codes). In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{TC}}$ is expanded accordingly \textcolor{red}{Owen: Add symbols for 4-0 resonance, and 1-3 resonances and 2-2 resonance}:
\begin{align}
\mathcal{C}_{\rm{TC}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include rotational symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

@@

We explicitly demonstrate how the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the $U(1)$ topological winding number $\Phi_X$ histogram. This is shown in \figref{fig:u1_wind_qdpm}, where the different colors represent calculations that were initialized to different winding sectors. We confirm that $\Phi_{X,Y}$ are conserved modulo $3$, leading to $9$-fold topological degeneracy. Notice that $\Phi_X = \pm L/2$ correspond to ordered configurations, and therefore the probability of $\Phi_{XY}$ reaching these extrema vanishes in the RK state.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{u1_wind_qdpm.pdf}
    \caption{The histogram of the $U(1)$ winding number $\Phi_X$ as measured in the QDPM on a $16\times16$ lattice. Different colors correspond to calculations initialized to different winding sectors. Note that $\Phi_X$ is conserved modulo $3$, resulting in a $9$-fold topological degeneracy.}
    \label{fig:u1_wind_qdpm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}
        \centering
        \includegraphics[width=0.75\columnwidth]{dimer_gry_scale_qdpm.pdf}
        \caption{Gray scale of the dimer correlation function between the dimer in the upper
        left corner (colored red) and all other dimers on a $16\times16$ lattice. The 
         red dimers indicate correlations that are off the scale shown at the
        bottom of the figure. \textcolor{red}{Owen: make the origin dimer a different color that the off-scale dimers}.}
        \label{fig:dimer_gry_scale_qdpm}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spacial_dmr_cor.pdf}
    \caption{Correlation function between parallel dimers one the same (blue) and different (red) 
    sub-lattice on an $L=32$ lattice. The lines represent least-squares fits to exponential decay. }
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
   \begin{figure}[]
        \centering
        \includegraphics[width=0.75\columnwidth]{qdpm_dmr_vis_HM_16x16.pdf}
        \caption{QDPM dimer structure factor $\gamma=0.5$.}
        \label{fig:spacial_dmr_z3vis_cor}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
